---
title: "BIOS_HW#1"
author: "Jared Garfinkel"
date: "9/13/2019"
output: html_document
---

#Problem 1

This is the solution for Problem 1

```{r load libraries}
library(tidyverse)
library(arsenal)
library(dplyr)
library(knitr)
```

##Section a

This is a solution for section a.

There is a top row of text that can be ignored for the purposes of creating a table.

```{r import doc}

exercise <- read.csv("C:\\Users\\jared\\OneDrive\\Fall 2019\\Biostatistics\\HW1\\BIOS-HW1\\Exercise.csv", skip = 1)

```

Change the labels.

```{r labels}

my_labels <- list(Age = "Age (yrs)", Gender = "Gender", Race = "Race", HTN = "Hypertensive", T2DM = "Type 2 Diabetes", Depression = "Depression", Smokes = "Smoker")
```

Clean the output.

```{r clean output}
my_controls <- tableby.control(
               total = FALSE,
               test = FALSE,  # No test p-values yet
               numeric.stats = c("meansd", "medianq1q3"),
               cat.stats = c("countpct"),
               stats.labels = list(
               meansd = "Mean (SD)",
               medianq1q3 = "Median (Q1, Q3)",
               countpct = "N (%)"))
```

I found a function called mutate from recitation that is useful for showing categories. No idea why it works.

```{r categories}

table1 <- exercise %>%
  mutate(Group = factor(Group, labels = c("Intervention","Control"))) %>%
  mutate(Gender = factor(Gender, labels = c("Male", "Female"))) %>%
  mutate(Race = factor(Race, labels = c("African American", "Hispanic", "Caucasian"))) %>%
  mutate(HTN = factor(HTN, labels = c("Yes","No"))) %>%
  mutate(T2DM = factor(T2DM, labels = c("Yes", "No"))) %>%
  mutate(Depression = factor(Depression, labels = c("Yes","No"))) %>%
  mutate(Smokes = factor(Smokes,labels = c("Yes","No")))
```

This is a first table.

```{r solution 1a}
solution1a <- tableby(Group ~ Age + Gender + Race + HTN + T2DM + Depression + Smokes, 
                      data = table1, control = my_controls)
```

This is a solution.

```{r output 1a}
summary(solution1a, title = "Demographics and Co-Morbidities", labelTranslations = my_labels,text = TRUE)

```

## Section 1b

This is a solution for Section 1b.

Start with the labels.

```{r labels2}

my_labels2 <- list(PRE = "Systolic Blood Pressure (mm Hg)", POST = " ",
                   PRE.1 = "Diastolic Blood Pressure (mm Hg)", POST.1 = " ",
                   PRE.2 = "Body Mass Index", POST.2 = " ",
                   PRE.3 = "HDL Cholesterol", POST.3 = " ",
                   PRE.4 = "LDL Cholesterol", POST.4 = " ",
                   PRE.5 = "Blood Glucose (mmol/L)", POST.5 = " ")
```


```{r categories2}
table2 <- exercise %>%
  mutate(Group = factor(Group, labels = c("Intervention", "Control")))
```

Clean the output.

```{r controls2}
my_controls2 <- tableby.control(
               total = FALSE,
               test = FALSE,  # No test p-values yet
               numeric.stats = c("meansd", "medianq1q3"),
               cat.stats = c("countpct"),
               stats.labels = list(
               meansd = "Mean (SD)",
               medianq1q3 = "Median (Q1, Q3)",
               countpct = "N (%)"))
```

Use only the variables we are looking at.

```{r table2}
solution1b <- tableby(Group ~ PRE + POST + PRE.1 + POST.1 + 
                        PRE.2 + POST.2 + PRE.3 + POST.3 + 
                        PRE.4 + POST.4 + PRE.5 + POST.5,
                      data = table2, control = my_controls2)
```

Make the table.

```{r solution2a}
summary(solution1b, title="Metabolic Parameters", labelTranslations = my_labels2, text=TRUE)
```

Create a dataframe to pull data for a table.

```{r df}
mp_df <- tibble(
  Sys_u = mean(pull(table2,PRE)),
  Sys_m = median(pull(table2,PRE)),
  Sys_sd = sd(pull(table2,PRE)),
  Sys_iqr = iqr(pull(table2,PRE)),
  Sys_u2 = mean(pull(table2,POST)),
  Sys_m2 = median(pull(table2,POST)),
  Sys_sd2 = sd(pull(table2,POST)),
  Sys_iqr2 = iqr(pull(table2,POST))
)
```

Create a dataframe to pull labels for rows.

```{r results = 'asis', eval = FALSE}
kable(solution1b, caption = "Metabolic Parameters")
```

- | Intervention |         | Control |     |    
- | ------------ | ------- | ------- | --- |
- | Baseline | 6-months | Baseline | 6-months
Systolic Blood Pressure (mm Hg) | `r round(pull(mp_df,Sys_u), digits = 2)` +/- `r round(pull(mp_df,Sys_sd), digits = 2)` | `r round(pull(mp_df,Sys_u2), digits = 2)`+/- `r round(pull(mp_df,Sys_sd2), digits = 2)`