---
title: "BIOS_HW#1"
author: "Jared Garfinkel"
date: "9/13/2019"
tables: yes
output: html_document
---

#Problem 1

This is the solution for Problem 1

```{r load libraries}
library(tidyverse)
library(arsenal)
library(dplyr)
library(knitr)
```

##Section a

This is a solution for section a.

There is a top row of text that can be ignored for the purposes of creating a table.

```{r import doc}

exercise <- read.csv("C:\\Users\\jared\\OneDrive\\Fall 2019\\Biostatistics\\HW1\\BIOS-HW1\\Exercise.csv", skip = 1)

```

Change the labels.

```{r labels}

my_labels <- list(Age = "Age (yrs)", Gender = "Gender", Race = "Race", HTN = "Hypertensive", T2DM = "Type 2 Diabetes", Depression = "Depression", Smokes = "Smoker")
```

Clean the output.

```{r clean output}
my_controls <- tableby.control(
               total = FALSE,
               test = FALSE,  # No test p-values yet
               numeric.stats = c("meansd", "medianq1q3"),
               cat.stats = c("countpct"),
               stats.labels = list(
               meansd = "Mean (SD)",
               medianq1q3 = "Median (Q1, Q3)",
               countpct = "N (%)"))
```

I found a function called mutate from recitation that is useful for showing categories. No idea why it works.

```{r categories}

table1 <- exercise %>%
  mutate(Group = factor(Group, labels = c("Intervention","Control"))) %>%
  mutate(Gender = factor(Gender, labels = c("Male", "Female"))) %>%
  mutate(Race = factor(Race, labels = c("African American", "Hispanic", "Caucasian"))) %>%
  mutate(HTN = factor(HTN, labels = c("Yes","No"))) %>%
  mutate(T2DM = factor(T2DM, labels = c("Yes", "No"))) %>%
  mutate(Depression = factor(Depression, labels = c("Yes","No"))) %>%
  mutate(Smokes = factor(Smokes,labels = c("Yes","No")))
```

This is a first table.

```{r solution 1a}
solution1a <- tableby(Group ~ Age + Gender + Race + HTN + T2DM + Depression + Smokes, 
                      data = table1, control = my_controls)
```

This is a solution.

```{r output 1a}
summary(solution1a, title = "Demographics and Co-Morbidities", labelTranslations = my_labels,text = TRUE)

```

## Section 1bi

This is a solution for Section 1b.

Start with the labels.

```{r labels2, echo = FALSE}

my_labels2 <- list(PRE = "Systolic Blood Pressure (mm Hg)", POST = " ",
                   PRE.1 = "Diastolic Blood Pressure (mm Hg)", POST.1 = " ",
                   PRE.2 = "Body Mass Index", POST.2 = " ",
                   PRE.3 = "HDL Cholesterol", POST.3 = " ",
                   PRE.4 = "LDL Cholesterol", POST.4 = " ",
                   PRE.5 = "Blood Glucose (mmol/L)", POST.5 = " ")
```


```{r categories2, echo = FALSE}
table2 <- exercise %>%
  mutate(Group = factor(Group, labels = c("Control", "Intervention")))
```

Clean the output.

```{r controls2, echo = FALSE}
my_controls2 <- tableby.control(
               total = FALSE,
               test = FALSE,  # No test p-values yet
               numeric.stats = c("meansd", "medianq1q3"),
               cat.stats = c("countpct"),
               stats.labels = list(
               meansd = "Mean (SD)",
               medianq1q3 = "Median (Q1, Q3)",
               countpct = "N (%)"))
```

Use only the variables we are looking at.

```{r table2, echo = FALSE}
solution1b <- tableby(Group ~ PRE + POST + PRE.1 + POST.1 + 
                        PRE.2 + POST.2 + PRE.3 + POST.3 + 
                        PRE.4 + POST.4 + PRE.5 + POST.5,
                      data = table2, control = my_controls2)
```

Make a table.

```{r solution2a, echo = FALSE, eval = FALSE}
summary(solution1b, title="Metabolic Parameters", labelTranslations = my_labels2, text=TRUE)
```

Create a dataframe that splits data by factor.

```{r split data by factor, echo = FALSE, eval = FALSE}

# exp <- factor(table2$Group)

df_good <- exercise %>% group_by(Group)

df_good %>% summarize(
  pre = mean(PRE),
  post = mean(POST),
  pre.1 = mean(PRE.1),
  post.1 = mean(POST.1),
  pre.2 = mean(PRE.2),
  post.2 = mean(POST.2),
  pre.3 = mean(PRE.3),
  post.3 = mean(POST.3),
  pre.4 = mean(PRE.4),
  post.4 = mean(POST.4),
  pre.5 = mean(PRE.5),
  post.5 = mean(POST.5),
)
```

```{r df, echo = FALSE, eval = FALSE}
mp_df <- df_good %>% tibble(
  Sys_u = mean(PRE), 
  Sys_sd = sd(PRE),
  Sys_u2 = mean(POST), 
  Sys_sd2 = sd(POST)
)
```

```{r df1, echo = FALSE, eval = FALSE}
groupsys1 <- tableby(Group ~ PRE, data=table2, control = my_controls2)
groupsys2 <- tableby(Group ~ POST, data=table2, control = my_controls2)
groupdia1 <- tableby(Group ~ PRE.1, data=table2, control = my_controls2)
groupdia2 <- tableby(Group ~ POST, data=table2, control = my_controls2)
groupbmi1 <- tableby(Group ~ PRE, data=table2, control = my_controls2)
groupbmi2 <- tableby(Group ~ POST, data=table2, control = my_controls2)
grouphdl1 <- tableby(Group ~ PRE, data=table2, control = my_controls2)
grouphdl2 <- tableby(Group ~ POST, data=table2, control = my_controls2)
groupldl1 <- tableby(Group ~ PRE, data=table2, control = my_controls2)
groupldl2 <- tableby(Group ~ POST, data=table2, control = my_controls2)
groupglu1 <- tableby(Group ~ PRE, data=table2, control = my_controls2)
groupglu2 <- tableby(Group ~ POST, data=table2, control = my_controls2)
```

```{r df2, echo = FALSE, eval = FALSE}
table1 <- summary(groupsys1, title="Systolic Blood Pressure", labelTranslations = my_labels2, text=TRUE)
table2 <- summary(groupsys2, title="Systolic Blood Pressure 6-months", labelTranslations = my_labels2, text=TRUE)
kable(list(table1, table2))
```

```{r Pipes}
new_df <- select(exercise,Group,PRE,POST,PRE.1,POST.1,PRE.2,POST.2,PRE.3,POST.3,PRE.4,POST.4,PRE.5,POST.5)
new_df1 <- filter(new_df, Group == 1)
new_df2 <- filter(new_df, Group == 0)
```

: Changes in Metabolic Paramaters

------ | Intervention |         | Control |     |    
------ | -------- | ------- | ------- | ------ |
------ | Baseline | 6-months | Baseline | 6-months
Systolic Blood Pressure (mm Hg) | `r round(mean(pull(new_df1,PRE)), digits = 2)` +/- `r round(sd(pull(new_df1,PRE)), digits = 2)` | `r round(mean(pull(new_df1,POST)), digits = 2)` +/- `r round(sd(pull(new_df1,POST)), digits = 2)` | `r round(mean(pull(new_df2,PRE)), digits = 2)` +/- `r round(sd(pull(new_df2,PRE)), digits = 2)` | `r round(mean(pull(new_df2,POST)), digits = 2)` +/- `r round(sd(pull(new_df2,POST)), digits = 2)`
------ | `r round(median(pull(new_df1,PRE)), digits = 2)` (`r q1q3(pull(new_df1,PRE))`) | `r round(median(pull(new_df1,POST)), digits=2)` (`r q1q3(pull(new_df1,POST))`) | `r round(median(pull(new_df2,PRE)), digits = 2)` (`r q1q3(pull(new_df2,PRE))`) | `r round(median(pull(new_df2,POST)), digits=2)` (`r q1q3(pull(new_df2,POST))`)
Diastolic Blood Pressure (mm Hg) | `r round(mean(pull(new_df1,PRE.1)), digits = 2)` +/- `r round(sd(pull(new_df1,PRE.1)), digits = 2)` | `r round(mean(pull(new_df1,POST.1)), digits = 2)` +/- `r round(sd(pull(new_df1,POST.1)), digits = 2)` | `r round(mean(pull(new_df2,PRE.1)), digits = 2)` +/- `r round(sd(pull(new_df2,PRE.1)), digits = 2)` | `r round(mean(pull(new_df2,POST.1)), digits = 2)` +/- `r round(sd(pull(new_df2,POST.1)), digits = 2)`
------ | `r round(median(pull(new_df1,PRE.1)), digits = 2)` (`r q1q3(pull(new_df1,PRE.1))`) | `r round(median(pull(new_df1,POST.1)), digits=2)` (`r q1q3(pull(new_df1,POST.1))`) | `r round(median(pull(new_df2,PRE.1)), digits = 2)` (`r q1q3(pull(new_df2,PRE.1))`) | `r round(median(pull(new_df2,POST.1)), digits=2)` (`r q1q3(pull(new_df2,POST.1))`)
Body Mass Index | `r round(mean(pull(new_df1,PRE.2)), digits = 2)` +/- `r round(sd(pull(new_df1,PRE.2)), digits = 2)` | `r round(mean(pull(new_df1,POST.2)), digits = 2)` +/- `r round(sd(pull(new_df1,POST.2)), digits = 2)` | `r round(mean(pull(new_df2,PRE.2)), digits = 2)` +/- `r round(sd(pull(new_df2,PRE.2)), digits = 2)` | `r round(mean(pull(new_df2,POST.2)), digits = 2)` +/- `r round(sd(pull(new_df2,POST.2)), digits = 2)`
------ | `r round(median(pull(new_df1,PRE.2)), digits = 2)` (`r q1q3(pull(new_df1,PRE.2))`) | `r round(median(pull(new_df1,POST.2)), digits=2)` (`r q1q3(pull(new_df1,POST.2))`) | `r round(median(pull(new_df2,PRE.2)), digits = 2)` (`r q1q3(pull(new_df2,PRE.2))`) | `r round(median(pull(new_df2,POST.2)), digits=2)` (`r q1q3(pull(new_df2,POST.2))`)
HDL Cholesterol (mg/dL) | `r round(mean(pull(new_df1,PRE.3)), digits = 2)` +/- `r round(sd(pull(new_df1,PRE.3)), digits = 2)` | `r round(mean(pull(new_df1,POST.3)), digits = 2)` +/- `r round(sd(pull(new_df1,POST.3)), digits = 2)` | `r round(mean(pull(new_df2,PRE.3)), digits = 2)` +/- `r round(sd(pull(new_df2,PRE.3)), digits = 2)` | `r round(mean(pull(new_df2,POST.3)), digits = 2)` +/- `r round(sd(pull(new_df2,POST.3)), digits = 2)`
------ | `r round(median(pull(new_df1,PRE.3)), digits = 2)` (`r q1q3(pull(new_df1,PRE.3))`) | `r round(median(pull(new_df1,POST.3)), digits=2)` (`r q1q3(pull(new_df1,POST.3))`) | `r round(median(pull(new_df2,PRE.3)), digits = 2)` (`r q1q3(pull(new_df2,PRE.3))`) | `r round(median(pull(new_df2,POST.3)), digits=2)` (`r q1q3(pull(new_df2,POST.3))`)
LDL Cholesterol (mg/dL) | `r round(mean(pull(new_df1,PRE.4)), digits = 2)` +/- `r round(sd(pull(new_df1,PRE.4)), digits = 2)` | `r round(mean(pull(new_df1,POST.4)), digits = 2)` +/- `r round(sd(pull(new_df1,POST.4)), digits = 2)` | `r round(mean(pull(new_df2,PRE.4)), digits = 2)` +/- `r round(sd(pull(new_df2,PRE.4)), digits = 2)` | `r round(mean(pull(new_df2,POST.4)), digits = 2)` +/- `r round(sd(pull(new_df2,POST.4)), digits = 2)`
------ | `r round(median(pull(new_df1,PRE.4)), digits = 2)` (`r q1q3(pull(new_df1,PRE.4))`) | `r round(median(pull(new_df1,POST.4)), digits=2)` (`r q1q3(pull(new_df1,POST.4))`) | `r round(median(pull(new_df2,PRE.4)), digits = 2)` (`r q1q3(pull(new_df2,PRE.4))`) | `r round(median(pull(new_df2,POST.4)), digits=2)` (`r q1q3(pull(new_df2,POST.4))`)
Blood Glucose Level (mmol/L) | `r round(mean(pull(new_df1,PRE.5)), digits = 2)` +/- `r round(sd(pull(new_df1,PRE.5)), digits = 2)` | `r round(mean(pull(new_df1,POST.5)), digits = 2)` +/- `r round(sd(pull(new_df1,POST.5)), digits = 2)` | `r round(mean(pull(new_df2,PRE.5)), digits = 2)` +/- `r round(sd(pull(new_df2,PRE.5)), digits = 2)` | `r round(mean(pull(new_df2,POST.5)), digits = 2)` +/- `r round(sd(pull(new_df2,POST.5)), digits = 2)`
------ | `r round(median(pull(new_df1,PRE.5)), digits = 2)` (`r q1q3(pull(new_df1,PRE.5))`) | `r round(median(pull(new_df1,POST.5)), digits=2)` (`r q1q3(pull(new_df1,POST.5))`) | `r round(median(pull(new_df2,PRE.5)), digits = 2)` (`r q1q3(pull(new_df2,PRE.5))`) | `r round(median(pull(new_df2,POST.5)), digits=2)` (`r q1q3(pull(new_df2,POST.5))`)

# Problem 1bii

Two boxplots of intervention groups.

```{r graphics}
library(ggplot2)
```

```{r boxplot BMI, echo = FALSE, eval = FALSE}
ggplot(select(exercise,Group,PRE.2), aes(x = Group, y = PRE.2)) +
  geom_boxplot() +
  scale_fill_manual(values = c("blue", "red"),
                    labels = expression("Baseline", "6-months")) +
  xlab("Intervention") +
  ylab("Body Mass Index") +
  theme(legend.position = "top", legend.title = element_blank())
```

```{r boxplot LDL, echo = FALSE, eval = FALSE}
ggplot(exercise, select(pull(exercise, subset(PRE.4, Group == '1')), 
                     pull(exercise, subset(POST.4, Group == '1')))) +
  geom_boxplot() +
  scale_fill_manual(values = c("blue", "red"),
                    labels = expression("Baseline", "6-months")) +
  xlab("Intervention") +
  ylab("LDL Cholesterol (mg/dL)") +
  theme(legend.position = "top", legend.title = element_blank())
```

```{r boxplot2 BMI, echo = FALSE, eval = FALSE}
par(mfrow=c(1,2))
with(subset(exercise, Group == '1'), boxplot(PRE.2, xlab = "Baseline",
                                             col = "blue", main = "Intervention"))
with(subset(exercise, Group == '1'), boxplot(POST.2, xlab = "6-months",
                                              col = "red", main = "Intervention"))
```

```{r boxplot3 BMI, echo = FALSE, eval = FALSE}

BMI_boxplot <- select(filter(exercise, Group == 1), PRE.2, POST.2)

my_labels3 <- list(PRE.2 = "Baseline", POST.2 = "6-months")

boxplot(BMI_boxplot,
        names = my_labels3, ylab = "Body Mass Index",
        col = c("blue", "red"), main = "Intervention")
```

```{r boxplot3 LDL, echo = FALSE, eval = FALSE}

LDL_boxplot <- select(filter(exercise, Group == 1), PRE.4, POST.4)

my_labels4 <- list(PRE.4 = "Baseline", POST.4 = "6-months")

boxplot(LDL_boxplot,
        names = my_labels4, ylab = "LDL Cholesterol (mg/dL)",
        col = c("blue", "red"), main = "Intervention")
```

```{r boxplot4 BMI+LDL}
#BMI_boxplot2 <- select(filter(exercise, Group == 0), PRE.2, POST.2)
BMI_df1a <- pull(filter(exercise, Group == 1), PRE.2)
BMI_df1b <- pull(filter(exercise, Group == 1), POST.2)
BMI_df0a <- pull(filter(exercise, Group == 0), PRE.2)
BMI_df0b <- pull(filter(exercise, Group == 0), POST.2)
BMI_good <- cbind(BMI_df1a, BMI_df1b, BMI_df0a, BMI_df0b)
BMI_labels <- list(BMI_df1a = "Baseline", BMI_df1b = "6-months", BMI_df0a = "Baseline", BMI_df0b = "6-months")

LDL_df1a <- pull(filter(exercise, Group == 1), PRE.4)
LDL_df1b <- pull(filter(exercise, Group == 1), POST.4)
LDL_df0a <- pull(filter(exercise, Group == 0), PRE.4)
LDL_df0b <- pull(filter(exercise, Group == 0), POST.4)
LDL_good <- cbind(LDL_df1a, LDL_df1b, LDL_df0a, LDL_df0b)
LDL_labels <- list(LDL_df1a = "Baseline", LDL_df1b = "6-months", LDL_df0a = "Baseline", LDL_df0b = "6-months")


par(mfrow = c(2,1), mar=c(4,4,2,2))
with(exercise,
boxplot(BMI_good,
        names = BMI_labels,
        xlab = "",
        ylab = "Body Mass Index", 
        col = c("blue","red"), 
        main = "Effects of exercise on BMI and LDL"))
with(exercise,
boxplot(LDL_good,
        names = LDL_labels,
        xlab = "Intervention vs. Control",
        ylab = "LDL Cholesterol (mg/dL)", 
        col = c("blue","red"), 
        main = ""))
```

```{r unused, echo = FALSE, eval = FALSE}
boxplot(BMI_boxplot2, 
  xlab = "",
  ylab = "",
  col = c("blue","red"),
  main = "")
with(LDL_boxplot2,
     boxplot(LDL_boxplot2, data = LDL_boxplot2,
             xlab = "",
             ylab = "LDL Cholesterol (mg/dL)", 
             col = c("blue","red"),
             main = ""))
```

## Section 1biii

These findings indicate that among the intervention and control

groups there were no changes in BMI and LDL. This might be because

of the matching technique used, where participants and non-participants

were matched by gender. A better method may be to match the participants

and non-participants by smoking status, or other health-related

indicators.


